Cooling Requirements for the Vertical Shear Instability in Protoplanetary Disks

### 2015-04-23 appendix.tex
notation change from epsilon to h

% In \S\ref{approx_gov} we made the replacement $D\to\Omega_k^2$ before
% eliminating variables to obtain a single equation for $\delta v_z$ in
% vertically isothermal disks, Eq. \ref{vertiso_gov}.  
% This procedure ignores the vertical dependence of
% $D=\kappa^2(z) - \sigma^2$. We show here that this has no serious 
% consequence for thin disks.

@@

\begin{align}\label{full_ode}
  0 =& \frac{d^2\delta v_z}{dz^2} + \left[1 + \frac{\ii k_x c_s^2
      q}{Dr} - \frac{k_x^2c_s^2}{\left(k_x^2c_s^2 + \chi
        D\right)}\frac{qc_s^2F}{Dr^2}\right]\frac{d\ln\rho}{dz}\frac{d\delta
    v_z}{dz} \notag\\
  &+ \left\{\sigma^2\left(\frac{k_x^2}{D} +
      \frac{\chi}{c_s^2}\right) + \left(\chi + \frac{\ii k_x c_s^2
        q}{Dr}\right)\frac{d^2\ln\rho}{dz^2} -
    \frac{c_s^2}{D}\left(\frac{d\ln\rho}{dz}\right)^2\left(k_x^2 -
      \frac{\ii k_x q}{r}\right)
   \left[\left(1-\chi\right) +
     \frac{\chi}{\left(k_x^2c_s^2 + \chi D\right)}\frac{qc_s^2 F}{r^2}\right] 
   \right\}\delta v_z,
\end{align}
where we recall $\chi = \left(1-\ii\sigma t_c\right)/\left(1-\ii\sigma
t_c \gamma\right)$. Making the low frequency and Keplerian
approximations is equivalent to the replacement $D\to
\Omega_k^2$. This gives, in terms of dimensionless variables,  
\begin{align}
  &\delta v_z ^{\prime\prime} + \left[1 + \ii\epsilon q\hat{k} -
    \frac{ \hat{k}^2}{
      \left(\hat{k}^2+\chi\right)}q \epsilon^2F\right]\ln\rho^{\prime}\delta v_z^\prime +
  \left\{\left(\chi + \ii \epsilon q
      \hat{k}\right)\ln\rho^{\prime\prime} - \ln\rho^{\prime
      2}\left(\khat^2 -
      \ii\epsilon
      q\hat{k}\right)\left[1 - \chi +
      \frac{\chi}{\left(\hat{k}^2+\chi\right)}q\epsilon^2F\right]\right\}\delta v_z \notag\\&=
  -\hat{\sigma}^2\left(\hat{k}^2+\chi\right)\delta v_z\label{adia_diso3}.
\end{align} 
Since we are interested in thin disks with $\epsilon \ll 1$ and $|q| =
O(1)$, we neglect the terms proportional to $\epsilon^2$, which gives 
Eq. \ref{vertiso_gov_nondim} (the dimensional version being
Eq. \ref{vertiso_gov}). This is equivalent to neglecting the   
vertical dependence of $D$ by setting $D\to\Omega_k^2$ in
Eq. \ref{ode_w} before eliminating variables in favor of $\delta
v_z$, which is the simpler way to obtain Eq. \ref{vertiso_gov}.

@@

%Eq. \ref{adia_diso3} differs from Eq. \ref{vertiso_gov_nondim} by terms
%proportional to $\epsilon^2$. For a thin disk ($\epsilon\ll1$) with
%$|q|=O(1)$, as considered in this work, neglecting these terms has no
%qualitative effect on the discussion in \S\ref{analytic_relax}. 
% In deriving Eq. \ref{vertiso_gov}, the vertical dependence of
% $D=\kappa^2(z)-\sigma^2$ was ignored because the replacement
% $D\to\Omega_k^2$ was made before eliminating variables in favor of
% $\delta v_z$. In Appendix \ref{adia_improve} we show that making this
% replacement after eliminating variables, which involves $dD/dz$,
% introduces terms of $O(\epsilon^2)$. Since we are interested in 
% thin disks ($\epsilon\ll 1 $), e.g. protoplanetary disks where 
% $\epsilon\lesssim 0.1$ \citep{chiang10}, these terms introduce
% unnecessary complexity for the present discussion.

@@

The background radial disk structure explicitly appears 
in the linearized mass, radial momentum and energy equations
(Eq. \ref{lin_mass}, \ref{lin_vx} and \ref{lin_energy}) as the terms
proportional to $\hat{g}_c$. Nominally $\hat{g}_c=1$.      
In the VGSB (and SSB) these terms are neglected by setting
$\hat{g}_c=0$. For a power-law disk, these radial gradients  
are $O(r^{-1})$, and they appear in comparison with terms of
$O(k_x)$. The neglected terms therefore have a relative magnitude of
$O(\epsilon/\khat)$, which is small for thin disks ($\epsilon\ll1$)
and/or small radial wavelengths ($\khat\gg 1$).

@@

\subsection{Spurious growth of adiabatic perturbations in the VGSB}\label{analytic_adia} 
A limitation of the VGSB is that it cannot be used for 
adiabatic flow, even if the above conditions are satisfied. We explain this by considering vertically isothermal
disks and setting  $\beta\to\infty$ in 
Eq. \ref{vertiso_gov_nondim} % (which itself is obtained by setting
% $\hat{g}_c=0$ in Eq. \ref{lin_mass}---\ref{lin_energy})
to give 
\begin{align}
  0 =\dd v_z^{\prime\prime} + \left(1 + \ii \epsilon q
    \hat{k}\right)\left(\ln\rho^{\prime}\delta v_z\right)^\prime
  +\left\{\hat{\sigma}^2\left(\frac{1}{\gamma}+\hat{k}^2\right) 
    -\left(\frac{\gamma-1}{\gamma}\right)\left[\ln\rho^{\prime\prime}+\hat{k}^2\left(1-\frac{\ii\epsilon  
          q}{\hat{k}}\right)\ln\rho^{\prime 2}\right]\right\}\delta v_z.\label{adia_iso3}
\end{align}
We multiply Eq. \ref{adia_iso3} by $\rho\delta v_z^*$ and
integrate vertically, assume boundary terms vanish when integrating by
parts, to obtain
\begin{align}
  \hat{\sigma}^2\left(\frac{1}{\gamma} +
    \hat{k}^2\right)\int_{\zhat_1}^{\zhat_2}\rho|\delta
  v_z|^2 d\zhat 
  =&  \left(\frac{\gamma-1}{\gamma}\right)
  \int_{\zhat_1}^{\zhat_2}\rho|\delta v_z^\prime|^2 d\zhat
  +\frac{1}{\gamma}\int_{\zhat_1}^{\zhat_2}\frac{1}{\rho}|(\rho\delta
  v_z)^\prime|^2 d\zhat\notag\\
&+
  \left(\frac{\gamma-1}{\gamma}\right)\hat{k}^2\left(1-\frac{\ii\epsilon
      q}{\hat{k}}\right) \int_{\zhat_1}^{\zhat_2}\rho\ln\rho^{\prime
    2}|\delta v_z|^2 d\zhat
+ \ii\epsilon q \hat{k}
  \int_{\zhat_1}^{\zhat_2}\ln\rho^\prime(\rho\delta v_z^*)^\prime
  \delta v_z d\zhat.\label{adia_integral}
\end{align}
% When $q\equiv0$, all the terms on the right-hand-side (RHS) are real. Then
% $\hat{\sigma}^2$ is real, and  $\hat{\sigma}^2>0$ if $\gamma>1$. As
% expected, a sub-adiabatically stratified disk is stable in the absence
% of vertical shear. 
In the presence of vertical shear $q\neq0$, Eq. \ref{adia_integral}
suggests $\hat{\sigma}^2$ is complex for real $\khat$, which implies the
possibility of instability, regardless of the value of
$|\gamma-1|$. This is contradicts the second Solberg-Hoiland criterion 
(Eq. \ref{solberg2}), which imply instability for adiabatic
perturbations in a thin, vertically isothermal disk would require 
$|\gamma-1|\ll1$. 
%also need real freq^2 in adia limit 
This is because the VGSB retains the global radial
disk structure responsible for vertical shear
($\p_z\Omega$), but ignores it elsewhere in the linear problem. 
Nevertheless, we demonstrate below
that this inconsistency is unimportant for the VSI because it occurs
for $\beta\ll1$.   %and small scale

@@

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 0.0cm 0cm
  0cm]{figures/gcorr_compare} 
  \caption{Growth rate of the fundamental VSI mode as a function of
    the thermal relaxation time $\beta$. The `Analytical (VGSB)' curve
    is calculated from the dispersion relation Eq. \ref{relax_disp};
    the `Numerical (VGSB)' curve is obtained by numerically solving
    the linear problem, Eq. \ref{lin_mass}---\ref{lin_energy}, with $\hat{g}_c=0$
    (which defines the linear VGSB equations); and the `Numerical
    (Full)' curve is obtained by numerically solving 
    Eq. \ref{lin_mass}---\ref{lin_energy}  with $\hat{g}_c=1$, which
    accounts for the radial disk structure.  
    The disk parameters are $(\gamma, \Gamma)=(1.4, 1.011)$ and
    $(p,q,\epsilon)=(-1.5,-1,0.05)$. The perturbation
    wavenumber is $\khat=30$. 
    \label{gcorr_compare}}  
\end{figure}

\section{Linear problem in the isothermal limit}\label{iso_discuss}  
Here we summarize selected results for isothermal
perturbations ($\beta\equiv 0$) in vertically isothermal disks 
($\Gamma=1$) in the VGSB framework. In this case Eq. \ref{ode_Q}
becomes    
\begin{align} 
  Q = W 
\end{align}
(i.e. $\delta P = c_s^2\delta \rho$)% , and so $  \ii\hat{\sigma}c_s\delta v_z = W^\prime $
% from Eq. \ref{lin_vz}
. For isothermal perturbations it is 
simpler to work with an equation for $W$ by substituting $Q=W$ into
Eq. \ref{ode_w} and using Eq. \ref{ode_vz} to eliminate $\delta v_z$. 
In this case, we shall not yet make the low frequency
approximation, but first make the  Keplerian approximation. We
obtain, in terms of dimensionless variables, 
\begin{align}
  0 = W^{\prime\prime} + \left[\ln\rho^\prime - \frac{\ii \epsilon q \hat{k} f(\zhat)
      }{1-\hat{\sigma}^2} \right]W^\prime +
  \hat{\sigma}^2\left(1+\frac{\hat{k}^2}{1-\hat{\sigma}^2}\right)W,\label{iso_ode0} 
\end{align}
where for discussion purposes we have defined $f(\zhat)$ such that
\begin{align}\label{fz_shear}
  \frac{d\Omega^2}{d\hat{z}} = \epsilon^2q f(\hat{z})\Omega_k^2.
\end{align}
By comparing Eq. \ref{fz_shear} with Eq. \ref{vertical_shear_ex}, we
see that $f(\zhat)= 
\zhat\left(1+\epsilon^2\zhat^2\right)^{-3/2}$. More generally, though,
$f$ may be regarded as a representation of the vertical
shear profile. Physically, we expect there is a maximum value of 
$|d\Omega/dz|$, the existence of which should limit the growth rate of
the VSI,  as remarked in \S\ref{vshear_def}. We explicitly demonstrate
this below.

\subsection{Stability conditions in the low-frequency limit}
In the low-frequency limit we assume $|\hat{\sigma}|\ll 1$ and
approximate Eq. \ref{iso_ode0} as 
\begin{align}
  0 = W^{\prime\prime} + \left[\ln\rho^\prime - \ii \epsilon q \hat{k}
    f(\zhat)\right]W^\prime +
  \hat{\sigma}^2\left(1+\hat{k}^2\right)W. \label{iso_ode1}
\end{align}
\subsubsection{Necessity of vertical shear for
  instability}\label{integral_relation} 
We can establish a necessary condition for instability by multiplying
Eq. \ref{iso_ode1} by $\rho W^*$ and integrate vertically from
$\zhat=\zhat_1$ to $z=\zhat_2$. We neglect boundary 
terms when integrating by parts, by assuming $W$ or
$W^\prime$ vanishes at the boundaries, or that the boundaries are 
sufficiently far away so that the boundary terms are negligible because of the
decaying background density with increasing height. Then,
\begin{align}
  \hat{\sigma}^2\left(1+\hat{k}^2\right)\int_{\zhat_1}^{\zhat_2}\rho|W|^2d\zhat
  =\int_{\zhat_1}^{\zhat_2}\rho|W^\prime|^2d\zhat 
  +\ii \epsilon q \hat{k}\int_{\zhat_1}^{\zhat_2}\rho f(\zhat) W^*W^\prime d\zhat.\label{integral_relation1}
\end{align}
It follows that for instability ($\imag\hat{\sigma}>0$), it is necessary to
have $q\neq0$ or more generally $d\Omega^2/dz\neq 0$, i.e. there must
be vertical shear. A similar conclusion can be reached in the
equivalent linear problem in global cylindrical disk geometry.

\subsubsection{Maximum growth rate}\label{max_growth1}
Here we show that the growth rate is limited by the maximum vertical
shear rate in the domain. % Writing $\hat{\sigma} = \hat{\omega} +
% \ii\hat{\nu}$ with real $\hat{\omega}$ and $\hat{\nu}$, 
The real and imaginary parts of 
Eq. \ref{integral_relation1} are
\begin{align}
  \left(\hat{\omega}^2-\hat{\nu}^2\right)\left(1+\hat{k}^2\right)
  \int_{\zhat_1}^{\zhat_2}\rho|W|^2d\hat{z} -
  \int_{\zhat_1}^{\zhat_2}\rho|W^\prime|^2d\hat{z}
  \notag
  =\real\left[\ii
    \epsilon\hat{k}q \int_{\zhat_1}^{\zhat_2}\rho
    f(\hat{z}) W^*W^\prime d\zhat\right], \\
   2\hat{\omega}\hat{\nu}\left(1+\hat{k}^2\right)
  \int_{\zhat_1}^{\zhat_2}\rho|W|^2d\hat{z}
  =\imag\left[\ii
    \epsilon\hat{k}q \int_{\zhat_1}^{\zhat_2}\rho
    f(\hat{z}) W^*W^\prime d\hat{z}\right],
\end{align}
where we recall $\hat{\omega} = \real \hat{\sigma}$ and
$\hat{\nu}=\imag\hat{\sigma}$. 
Adding the square of these equations give
\begin{align}
  \left[|\hat{\sigma}|^2\left(1+\hat{k}^2\right)
    \int_{\zhat_1}^{\zhat_2}\rho|W|^2d\hat{z} -
    \int_{\zhat_1}^{\zhat_2}\rho\left|W^\prime \right|^2d\hat{z}\right]^2
  +4\hat{\nu}^2\left(1+\hat{k}^2\right) 
  \int_{\zhat_1}^{\zhat_2}\rho
  |W|^2d\hat{z}\int_{\zhat_1}^{\zhat_2}\rho\left|W^\prime \right|^2d\hat{z}
  =\left|\ii
    \epsilon\hat{k}q\int_{\zhat_1}^{\zhat_2}\rho
    f(\hat{z}) W^*W^\prime d\hat{z}\right|^2.
\end{align}
It is clear that
\begin{align}\label{sigma_finite_domain} 
  4\hat{\nu}^2\left(1+\hat{k}^2\right) 
  \int_{\zhat_1}^{\zhat_2}\rho
  |W|^2d\hat{z}\int_{\zhat_1}^{\zhat_2}\rho\left|W^\prime
  \right|^2d\hat{z} 
  \leq\left|
    \epsilon\hat{k}q\int_{\zhat_1}^{\zhat_2}\rho
    f(\hat{z}) W^*W^\prime d\hat{z}\right|^2.
\end{align}
On the left hand side of this inequality, we apply the Cauchy-Schwarz
inequality to obtain
\begin{align}
  \left( \int_{\zhat_1}^{\zhat_2}\rho
    |W|\left|W^\prime \right|d\hat{z}\right)^2\leq
  \int_{\zhat_1}^{\zhat_2}\rho 
  |W|^2d\hat{z}\int_{\zhat_1}^{\zhat_2}\rho\left|W^\prime \right|^2d\hat{z}.
\end{align}
On the right hand side of Eq. \ref{sigma_finite_domain} we have
\begin{align}
  \left|\int_{\zhat_1}^{\zhat_2}\rho
    f(\hat{z}) W^*W^\prime d\hat{z}\right|\leq \int_{\zhat_1}^{\zhat_2}\rho
  \left|f(\hat{z})W^*W^\prime \right|d\hat{z}
  \leq
  \mathrm{max}\left(|f|\right)\int_{\zhat_1}^{\zhat_2}\rho
  |W|\left|W^\prime \right|d\hat{z},
\end{align}
where $\mathrm{max}(|f|)$ is the maximum value of $|f|$ in
$\zhat\in[\zhat_1,\zhat_2]$. Inserting these inequalities into
Eq. \ref{sigma_finite_domain} gives
\begin{align}\label{max_growth}
  |\hat{\nu}|\leq
  \frac{\epsilon |\hat{k} q|}{2\sqrt{1+\hat{k}^2}}\mathrm{max}(|f|) < \frac{\epsilon |q|}{2}\mathrm{max}(|f|) . 
\end{align}
%Recalling that $f$ represents vertical shear (Eq. \ref{fz_shear}), 
It
follows that the maximum possible growth rate of unstable modes,
satisfying the above boundary conditions, is limited by the maximum
vertical shear rate in the domain considered,
\begin{align}
  \nu < \mathrm{max} \left|r\frac{d\Omega}{dz}\right|, 
\end{align}
as expected on physical grounds.

In practice, one might consider a vertical domain of a few scale 
heights in a thin disk with $|q|=O(1)$. In this case, $|\epsilon\zhat|\ll1$ and  
$f\simeq \zhat$, so that $\mathrm{max}|f| = O(1)$, implying a
maximum growth rate $O(\epsilon \Omega_k)$, consistent with numerical
results.

% For vertical domains smaller 
% than this height, the growth rate is limited by $|f|$ at the vertical
% boundaries considered. For $|\epsilon\hat{z}|\ll1$ we have $f\simeq
% \hat{z}$, in which case we can expect growth rates to increase at most linearly
% with respect to the vertical domain size.   %no faster than linear

>>>>

% In \S\ref{approx_gov} we made the replacement $D\to\Omega_K^2$ before
% eliminating variables to obtain a single equation for $\delta v_z$ in
% vertically isothermal disks, Eq. \ref{vertiso_gov}.  
% This procedure ignores the vertical dependence of
% $D=\kappa^2(z) - \sigma^2$. We show here that this has no serious 
% consequence for thin disks.

@@

\begin{align}\label{full_ode}
  0 =& \frac{d^2\delta v_z}{dz^2} + \left[1 + \frac{\ii k_x c_s^2
      q}{Dr} - \frac{k_x^2c_s^2}{\left(k_x^2c_s^2 + \chi
        D\right)}\frac{qc_s^2F}{Dr^2}\right]\frac{d\ln\rho}{dz}\frac{d\delta
    v_z}{dz} \notag\\
  &+ \left\{\sigma^2\left(\frac{k_x^2}{D} +
      \frac{\chi}{c_s^2}\right) + \left(\chi + \frac{\ii k_x c_s^2
        q}{Dr}\right)\frac{d^2\ln\rho}{dz^2} -
    \frac{c_s^2}{D}\left(\frac{d\ln\rho}{dz}\right)^2\left(k_x^2 -
      \frac{\ii k_x q}{r}\right)
   \left[\left(1-\chi\right) +
     \frac{\chi}{\left(k_x^2c_s^2 + \chi D\right)}\frac{qc_s^2 F}{r^2}\right] 
   \right\}\delta v_z,
\end{align}
where we recall $\chi = \left(1-\ii\sigma t_c\right)/\left(1-\ii\sigma
t_c \gamma\right)$. Making the low frequency and Keplerian
approximations is equivalent to the replacement $D\to
\Omega_K^2$. This gives, in terms of dimensionless variables,  
\begin{align}
  &\delta v_z ^{\prime\prime} + \left[1 + \ii h q\hat{k} -
    \frac{ \hat{k}^2}{
      \left(\hat{k}^2+\chi\right)}q h^2F\right]\ln\rho^{\prime}\delta v_z^\prime +
  \left\{\left(\chi + \ii h q
      \hat{k}\right)\ln\rho^{\prime\prime} - \ln\rho^{\prime
      2}\left(\khat^2 -
      \ii h
      q\hat{k}\right)\left[1 - \chi +
      \frac{\chi}{\left(\hat{k}^2+\chi\right)}qh^2F\right]\right\}\delta v_z \notag\\&=
  -\hat{\sigma}^2\left(\hat{k}^2+\chi\right)\delta v_z\label{adia_diso3}.
\end{align} 
Since we are interested in thin disks with $h \ll 1$ and $|q| =
O(1)$, we neglect the terms proportional to $h^2$, which gives 
Eq. \ref{vertiso_gov_nondim} (the dimensional version being
Eq. \ref{vertiso_gov}). This is equivalent to neglecting the   
vertical dependence of $D$ by setting $D\to\Omega_K^2$ in
Eq. \ref{ode_w} before eliminating variables in favor of $\delta
v_z$, which is the simpler way to obtain Eq. \ref{vertiso_gov}.

@@

%Eq. \ref{adia_diso3} differs from Eq. \ref{vertiso_gov_nondim} by terms
%proportional to $h^2$. For a thin disk ($h\ll1$) with
%$|q|=O(1)$, as considered in this work, neglecting these terms has no
%qualitative effect on the discussion in \S\ref{analytic_relax}. 
% In deriving Eq. \ref{vertiso_gov}, the vertical dependence of
% $D=\kappa^2(z)-\sigma^2$ was ignored because the replacement
% $D\to\Omega_K^2$ was made before eliminating variables in favor of
% $\delta v_z$. In Appendix \ref{adia_improve} we show that making this
% replacement after eliminating variables, which involves $dD/dz$,
% introduces terms of $O(h^2)$. Since we are interested in 
% thin disks ($h\ll 1 $), e.g. protoplanetary disks where 
% $h\lesssim 0.1$ \citep{chiang10}, these terms introduce
% unnecessary complexity for the present discussion.

@@

The background radial disk structure explicitly appears 
in the linearized mass, radial momentum and energy equations
(Eq. \ref{lin_mass}, \ref{lin_vx} and \ref{lin_energy}) as the terms
proportional to $\hat{g}_c$. Nominally $\hat{g}_c=1$.      
In the VGSB (and SSB) these terms are neglected by setting
$\hat{g}_c=0$. For a power-law disk, these radial gradients  
are $O(r^{-1})$, and they appear in comparison with terms of
$O(k_x)$. The neglected terms therefore have a relative magnitude of
$O(h/\khat)$, which is small for thin disks ($h\ll1$)
and/or small radial wavelengths ($\khat\gg 1$).

@@

\subsection{Spurious growth of adiabatic perturbations in the VGSB}\label{analytic_adia} 
A limitation of the VGSB is that it cannot be used for 
adiabatic flow, even if the above conditions are satisfied. We explain this by considering vertically isothermal
disks and setting  $\beta\to\infty$ in 
Eq. \ref{vertiso_gov_nondim} % (which itself is obtained by setting
% $\hat{g}_c=0$ in Eq. \ref{lin_mass}---\ref{lin_energy})
to give 
\begin{align}
  0 =\dd v_z^{\prime\prime} + \left(1 + \ii h q
    \hat{k}\right)\left(\ln\rho^{\prime}\delta v_z\right)^\prime
  +\left\{\hat{\sigma}^2\left(\frac{1}{\gamma}+\hat{k}^2\right) 
    -\left(\frac{\gamma-1}{\gamma}\right)\left[\ln\rho^{\prime\prime}+\hat{k}^2\left(1-\frac{\ii h  
          q}{\hat{k}}\right)\ln\rho^{\prime 2}\right]\right\}\delta v_z.\label{adia_iso3}
\end{align}
We multiply Eq. \ref{adia_iso3} by $\rho\delta v_z^*$ and
integrate vertically, assume boundary terms vanish when integrating by
parts, to obtain
\begin{align}
  \hat{\sigma}^2\left(\frac{1}{\gamma} +
    \hat{k}^2\right)\int_{\zhat_1}^{\zhat_2}\rho|\delta
  v_z|^2 d\zhat 
  =&  \left(\frac{\gamma-1}{\gamma}\right)
  \int_{\zhat_1}^{\zhat_2}\rho|\delta v_z^\prime|^2 d\zhat
  +\frac{1}{\gamma}\int_{\zhat_1}^{\zhat_2}\frac{1}{\rho}|(\rho\delta
  v_z)^\prime|^2 d\zhat\notag\\
&+
  \left(\frac{\gamma-1}{\gamma}\right)\hat{k}^2\left(1-\frac{\ii h
      q}{\hat{k}}\right) \int_{\zhat_1}^{\zhat_2}\rho\ln\rho^{\prime
    2}|\delta v_z|^2 d\zhat
+ \ii h q \hat{k}
  \int_{\zhat_1}^{\zhat_2}\ln\rho^\prime(\rho\delta v_z^*)^\prime
  \delta v_z d\zhat.\label{adia_integral}
\end{align}
% When $q\equiv0$, all the terms on the right-hand-side (RHS) are real. Then
% $\hat{\sigma}^2$ is real, and  $\hat{\sigma}^2>0$ if $\gamma>1$. As
% expected, a sub-adiabatically stratified disk is stable in the absence
% of vertical shear. 
In the presence of vertical shear $q\neq0$, Eq. \ref{adia_integral}
suggests $\hat{\sigma}^2$ is complex for real $\khat$, which implies the
possibility of instability, regardless of the value of
$|\gamma-1|$. This is contradicts the second Solberg-Hoiland criterion 
(Eq. \ref{solberg2}), which imply instability for adiabatic
perturbations in a thin, vertically isothermal disk would require 
$|\gamma-1|\ll1$. 
%also need real freq^2 in adia limit 
This is because the VGSB retains the global radial
disk structure responsible for vertical shear
($\p_z\Omega$), but ignores it elsewhere in the linear problem. 
Nevertheless, we demonstrate below
that this inconsistency is unimportant for the VSI because it occurs
for $\beta\ll1$.   %and small scale

@@

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 0.0cm 0cm
  0cm]{figures/gcorr_compare} 
  \caption{Growth rate of the fundamental VSI mode as a function of
    the thermal relaxation time $\beta$. The `Analytical (VGSB)' curve
    is calculated from the dispersion relation Eq. \ref{relax_disp};
    the `Numerical (VGSB)' curve is obtained by numerically solving
    the linear problem, Eq. \ref{lin_mass}---\ref{lin_energy}, with $\hat{g}_c=0$
    (which defines the linear VGSB equations); and the `Numerical
    (Full)' curve is obtained by numerically solving 
    Eq. \ref{lin_mass}---\ref{lin_energy}  with $\hat{g}_c=1$, which
    accounts for the radial disk structure.  
    The disk parameters are $(\gamma, \Gamma)=(1.4, 1.011)$ and
    $(p,q,h)=(-1.5,-1,0.05)$. The perturbation
    wavenumber is $\khat=30$. 
    \label{gcorr_compare}}  
\end{figure}

\section{Linear problem in the isothermal limit}\label{iso_discuss}  
Here we summarize selected results for isothermal
perturbations ($\beta\equiv 0$) in vertically isothermal disks 
($\Gamma=1$) in the VGSB framework. In this case Eq. \ref{ode_Q}
becomes    
\begin{align} 
  Q = W 
\end{align}
(i.e. $\delta P = c_s^2\delta \rho$)% , and so $  \ii\hat{\sigma}c_s\delta v_z = W^\prime $
% from Eq. \ref{lin_vz}
. For isothermal perturbations it is 
simpler to work with an equation for $W$ by substituting $Q=W$ into
Eq. \ref{ode_w} and using Eq. \ref{ode_vz} to eliminate $\delta v_z$. 
In this case, we shall not yet make the low frequency
approximation, but first make the  Keplerian approximation. We
obtain, in terms of dimensionless variables, 
\begin{align}
  0 = W^{\prime\prime} + \left[\ln\rho^\prime - \frac{\ii h q \hat{k} f(\zhat)
      }{1-\hat{\sigma}^2} \right]W^\prime +
  \hat{\sigma}^2\left(1+\frac{\hat{k}^2}{1-\hat{\sigma}^2}\right)W,\label{iso_ode0} 
\end{align}
where for discussion purposes we have defined $f(\zhat)$ such that
\begin{align}\label{fz_shear}
  \frac{d\Omega^2}{d\hat{z}} = h^2q f(\hat{z})\Omega_K^2.
\end{align}
By comparing Eq. \ref{fz_shear} with Eq. \ref{vertical_shear_ex}, we
see that $f(\zhat)= 
\zhat\left(1+h^2\zhat^2\right)^{-3/2}$. More generally, though,
$f$ may be regarded as a representation of the vertical
shear profile. Physically, we expect there is a maximum value of 
$|d\Omega/dz|$, the existence of which should limit the growth rate of
the VSI,  as remarked in \S\ref{vshear_def}. We explicitly demonstrate
this below.

\subsection{Stability conditions in the low-frequency limit}
In the low-frequency limit we assume $|\hat{\sigma}|\ll 1$ and
approximate Eq. \ref{iso_ode0} as 
\begin{align}
  0 = W^{\prime\prime} + \left[\ln\rho^\prime - \ii h q \hat{k}
    f(\zhat)\right]W^\prime +
  \hat{\sigma}^2\left(1+\hat{k}^2\right)W. \label{iso_ode1}
\end{align}
\subsubsection{Necessity of vertical shear for
  instability}\label{integral_relation} 
We can establish a necessary condition for instability by multiplying
Eq. \ref{iso_ode1} by $\rho W^*$ and integrate vertically from
$\zhat=\zhat_1$ to $z=\zhat_2$. We neglect boundary 
terms when integrating by parts, by assuming $W$ or
$W^\prime$ vanishes at the boundaries, or that the boundaries are 
sufficiently far away so that the boundary terms are negligible because of the
decaying background density with increasing height. Then,
\begin{align}
  \hat{\sigma}^2\left(1+\hat{k}^2\right)\int_{\zhat_1}^{\zhat_2}\rho|W|^2d\zhat
  =\int_{\zhat_1}^{\zhat_2}\rho|W^\prime|^2d\zhat 
  +\ii h q \hat{k}\int_{\zhat_1}^{\zhat_2}\rho f(\zhat) W^*W^\prime d\zhat.\label{integral_relation1}
\end{align}
It follows that for instability ($\imag\hat{\sigma}>0$), it is necessary to
have $q\neq0$ or more generally $d\Omega^2/dz\neq 0$, i.e. there must
be vertical shear. A similar conclusion can be reached in the
equivalent linear problem in global cylindrical disk geometry.

\subsubsection{Maximum growth rate}\label{max_growth1}
Here we show that the growth rate is limited by the maximum vertical
shear rate in the domain. % Writing $\hat{\sigma} = \hat{\omega} +
% \ii\hat{\nu}$ with real $\hat{\omega}$ and $\hat{\nu}$, 
The real and imaginary parts of 
Eq. \ref{integral_relation1} are
\begin{align}
  \left(\hat{\omega}^2-\hat{\nu}^2\right)\left(1+\hat{k}^2\right)
  \int_{\zhat_1}^{\zhat_2}\rho|W|^2d\hat{z} -
  \int_{\zhat_1}^{\zhat_2}\rho|W^\prime|^2d\hat{z}
  \notag
  =\real\left[\ii
    h\hat{k}q \int_{\zhat_1}^{\zhat_2}\rho
    f(\hat{z}) W^*W^\prime d\zhat\right], \\
   2\hat{\omega}\hat{\nu}\left(1+\hat{k}^2\right)
  \int_{\zhat_1}^{\zhat_2}\rho|W|^2d\hat{z}
  =\imag\left[\ii
    h\hat{k}q \int_{\zhat_1}^{\zhat_2}\rho
    f(\hat{z}) W^*W^\prime d\hat{z}\right],
\end{align}
where we recall $\hat{\omega} = \real \hat{\sigma}$ and
$\hat{\nu}=\imag\hat{\sigma}$. 
Adding the square of these equations give
\begin{align}
  \left[|\hat{\sigma}|^2\left(1+\hat{k}^2\right)
    \int_{\zhat_1}^{\zhat_2}\rho|W|^2d\hat{z} -
    \int_{\zhat_1}^{\zhat_2}\rho\left|W^\prime \right|^2d\hat{z}\right]^2
  +4\hat{\nu}^2\left(1+\hat{k}^2\right) 
  \int_{\zhat_1}^{\zhat_2}\rho
  |W|^2d\hat{z}\int_{\zhat_1}^{\zhat_2}\rho\left|W^\prime \right|^2d\hat{z}
  =\left|\ii
    h\hat{k}q\int_{\zhat_1}^{\zhat_2}\rho
    f(\hat{z}) W^*W^\prime d\hat{z}\right|^2.
\end{align}
It is clear that
\begin{align}\label{sigma_finite_domain} 
  4\hat{\nu}^2\left(1+\hat{k}^2\right) 
  \int_{\zhat_1}^{\zhat_2}\rho
  |W|^2d\hat{z}\int_{\zhat_1}^{\zhat_2}\rho\left|W^\prime
  \right|^2d\hat{z} 
  \leq\left|
    h\hat{k}q\int_{\zhat_1}^{\zhat_2}\rho
    f(\hat{z}) W^*W^\prime d\hat{z}\right|^2.
\end{align}
On the left hand side of this inequality, we apply the Cauchy-Schwarz
inequality to obtain
\begin{align}
  \left( \int_{\zhat_1}^{\zhat_2}\rho
    |W|\left|W^\prime \right|d\hat{z}\right)^2\leq
  \int_{\zhat_1}^{\zhat_2}\rho 
  |W|^2d\hat{z}\int_{\zhat_1}^{\zhat_2}\rho\left|W^\prime \right|^2d\hat{z}.
\end{align}
On the right hand side of Eq. \ref{sigma_finite_domain} we have
\begin{align}
  \left|\int_{\zhat_1}^{\zhat_2}\rho
    f(\hat{z}) W^*W^\prime d\hat{z}\right|\leq \int_{\zhat_1}^{\zhat_2}\rho
  \left|f(\hat{z})W^*W^\prime \right|d\hat{z}
  \leq
  \mathrm{max}\left(|f|\right)\int_{\zhat_1}^{\zhat_2}\rho
  |W|\left|W^\prime \right|d\hat{z},
\end{align}
where $\mathrm{max}(|f|)$ is the maximum value of $|f|$ in
$\zhat\in[\zhat_1,\zhat_2]$. Inserting these inequalities into
Eq. \ref{sigma_finite_domain} gives
\begin{align}\label{max_growth}
  |\hat{\nu}|\leq
  \frac{h |\hat{k} q|}{2\sqrt{1+\hat{k}^2}}\mathrm{max}(|f|) < \frac{h |q|}{2}\mathrm{max}(|f|) . 
\end{align}
%Recalling that $f$ represents vertical shear (Eq. \ref{fz_shear}), 
It
follows that the maximum possible growth rate of unstable modes,
satisfying the above boundary conditions, is limited by the maximum
vertical shear rate in the domain considered,
\begin{align}
  \nu < \mathrm{max} \left|r\frac{d\Omega}{dz}\right|, 
\end{align}
as expected on physical grounds.

In practice, one might consider a vertical domain of a few scale 
heights in a thin disk with $|q|=O(1)$. In this case, $|h\zhat|\ll1$ and  
$f\simeq \zhat$, so that $\mathrm{max}|f| = O(1)$, implying a
maximum growth rate $O(h \Omega_K)$, consistent with numerical
results.

% For vertical domains smaller 
% than this height, the growth rate is limited by $|f|$ at the vertical
% boundaries considered. For $|h\hat{z}|\ll1$ we have $f\simeq
% \hat{z}$, in which case we can expect growth rates to increase at most linearly
% with respect to the vertical domain size.   %no faster than linear
